An Open Platform for Managing Short Programming Exercises

### 2013-06-23 paper/paper.tex
edits after a proofreading pass

We describe CloudCoder, an open platform for creating, assigning, and sharing
short programming exercises for a variety of languages (currently C/C++, Java,
Python and Ruby).  Like other similar systems, CloudCoder is web-based,
letting students write code directly in a web browser, click the
``submit'' button, and receive immediate feedback.  Unlike other
systems, which tend to be closed, or commercial, or both, CloudCoder
is a completely open platform.  The code for the system is
open-source, and exercises written for CloudCoder are shared to a
central repository under a permissive Creative Commons license
(BY-SA).  Finally, CloudCoder collects detailed data that faculty can use for
educational research.

@@

Recent years have seen an explosion of online services for learning
to program:  Udacity\cite{udacity}, Codecademy\cite{codeacademy},
ProgZoo\cite{progzoo}, 
Codingbat\cite{codingbat}, Coursera\cite{coursera}, CodeLab\cite{turingscraft},
Problets\cite{Kumar:2005:GPA:1163405.1163408}, CS Circles\cite{Pritchard:2013:CCI:2445196.2445370} and
several others.  These systems are very powerful, and show great promise to
improve introductory computer science.  Given all of these
very good systems, one may ask why we thought
%it a good career move
%to build yet another system.
the world needed another such system.

@@

Our primary motivation in developing CloudCoder
was {\em freedom}: we wanted a system where
the software and exercises could be freely redistributed, and
were not under the control of any central authority.  This motivation
is not entirely philosophical: we felt that locking down either
the software or the exercises would result in a barrier to adoption,
limiting the potential impact of the system.
As we will discuss in Section~\ref{sec:questions}, we think that there
are some interesting research questions that can be addressed
best through the existence of a ``free like air'' programming
exercise system: for example, we are very interested in finding out
whether exercises can be reused effectively in different courses
at different institutions.

@@

Figure~\ref{screenshot} shows a screenshot of a CloudCoder exercise.
CloudCoder uses the Ace editor\cite{ace}, which supports
syntax highlighting and automatic indentation for most programming languages.
The submit button runs the code against a series of
test cases. Once the test results are received, they are presented
directly in the browser, giving the student feedback on the
correctness of the submission.
CloudCoder captures the code's standard output and
standard error streams to help students with debugging.

@@

CloudCoder supports two types of exercises:  {\em method} or {\em function-based}
exercises, and {\em whole-program} exercises.  Function based exercises are
very similar to CodingBat exercises: the student is required to write
a function/method that takes some number of parameter values,
does a computation, and returns a result value.
Students are given a description of
what the method should do, and a couple of example inputs with
expected results.  Writing a new function-based exercise
requires a description, some skeleton code (typically the method
header, although exercises can provide as much or as little code as
desired), and a series of inputs and their corresponding expected results.
Test cases can be marked as ``secret'' so that students don't see their
outputs, to discourage students from writing a special case to
recognize each input.

Whole-program exercises read input data from
the standard input stream, perform a computation, and write output to the standard output stream.
Expected output is checked by matching a regular expression against the
program's standard output stream.
Whole-program exercises are useful for two main reasons.  First,
they can be assigned early in the course, before functions/methods
have been introduced.  Second, they give the exercise author a
way of checking expected output that is more powerful than matching a return
value against a literal.  The main disadvantage is devising a
regular expression that is precise enough to distinguish correct output
from incorrect output, but forgiving enough to account for minor deviations
in whitespace, capitalization, and formatting of output.

@@

CloudCoder executes Java, Python, and Ruby code relatively securely using
a Java Security Manager.
(CloudCoder uses Jython\cite{jython} and JRuby\cite{jruby}
to execute Python and Ruby submissions, allowing us to take advantage of
the Java Security Manager for those languages.)
CloudCoder's execution of C/C++ code is somewhat less secure, although 
we hope to take advantage of a better sandboxing mechanism
in the near future, possibly based on the work of
David Malan\cite{Malan:2013:CSS:2445196.2445242} for Harvard's
online CS50 course.  However, because thus far the only users of
CloudCoder have been instructors and students,
the weak sandboxing has not been a significant concern.

@@

As the student works on an exercises, CloudCoder logs all of the
student's edits, and stores all submissions and test outcomes.
We hope that the combination of exercises and data will give faculty
and researchers
insights into the difficulty level of different exercises, keep a
record of common error patterns, and enable the development of
innovative data analyses.

One question is how to manage the data collected.  
Recent work on web-scale data collection by the
BlueJ team\cite{Utting:2012:WDG:2361276.2361278} lays out a number of important
questions about programming data, such as who should have access to the data, and what
mechanism should be used to access the data.  Sanders
et.\ al.\cite{Sanders:2008:DSE:1404520.1404534} reviewed a number of
the challenges involved in sharing empirical education data.

@@

Anonymized data that has been cleared by the proper Institutional
Review Boards is contributed to the Datashop, which
makes the data available to other researchers under whatever
restrictions were put in place by the overseeing IRBs.  The Datashop also
provides a suite of analysis and visualization tools for better making
sense of the ITS data.

@@

We recognize that ease of installation is a critical factor for
adoption of a new pedagogical technology for busy
faculty who may only have a few minutes to evaluate a new system.  We
are committed to making the installation process as simple and painless as 
possible.

@@

To date, each institution that has adopted CloudCoder in production has
provided Linux servers that we configured for them.  This has made
adoption very easy, but is not a model that is likely to scale well
to larger numbers of institutions.  We recently received a grant
from Amazon of credits for Amazon Web Services\cite{aws},
which we plan to use to directly host CloudCoder for interested
institutions.  We are also investigating easier ways for
users to self-host, such as enabling CloudCoder to run on
hosting services such as Google App Engine\cite{gae}.
%  We have asked Amazon for Amazon Web Services credits to
%provide CloudCoder ``in the cloud'' to adopting institutions, and we
%hope to add support for Google App Engine this summer.  Ease of
%adoption is critical for CloudCoder.

@@

{\bf How can programming exercises be integrated most effectively into a course?}
Programming exercises can be used for many purposes: self-test questions
to accompany readings and video lectures, in-class assessment (possibly
in the context of peer instruction), extra practice for students who
are struggling.  We would like to have specific, evidence-based recommendations
that we can give to instructors who are considering using programming
exercises in their courses.

>>>>

We describe CloudCoder, an open platform for creating, assigning, and sharing
short programming exercises for a variety of languages (currently C/C++, Java,
Python and Ruby).  Like other similar systems, CloudCoder is web-based,
letting students write code directly in a web browser, click the
``submit'' button, and receive immediate feedback.  Unlike other
systems, which tend to be closed, or commercial, or both, CloudCoder
is a completely open platform.  The code for the system is
open-source, and exercises written for CloudCoder may be shared to a
central repository under permissive licenses such as Creative Commons BY-SA.
Finally, CloudCoder collects detailed data that faculty can use for
educational research.

@@

Recent years have seen an explosion of online services for learning
to program:  Udacity\cite{udacity}, Codecademy\cite{codeacademy},
ProgZoo\cite{progzoo}, 
Codingbat\cite{codingbat}, Coursera\cite{coursera}, CodeLab\cite{turingscraft},
Problets\cite{Kumar:2005:GPA:1163405.1163408}, CS Circles\cite{Pritchard:2013:CCI:2445196.2445370} and
many others.  These systems are powerful, and show great promise to
improve introductory computer science.  Given the existence of all of these
systems, one may ask why we thought
%it a good career move
%to build yet another system.
the world needed another such system.

@@

Our primary motivation in developing CloudCoder
was {\em freedom}: we wanted a system where
the software and exercises could be freely redistributed, and
were not under the control of any central authority.  This motivation
is not entirely philosophical: we felt that locking down either
the software or the exercises would result in a barrier to adoption,
limiting the potential impact of the system.
As we will discuss in Section~\ref{sec:questions}, we think that there
are some interesting research questions that can be addressed
best through the existence of a ``free like air'' programming
exercise system: for example, we are interested in finding out
whether exercises can be reused effectively in different courses
at different institutions.

@@

Figure~\ref{screenshot} shows a screenshot of a CloudCoder exercise.
CloudCoder uses Ace\cite{ace}, an open source web-based
programmer's editor, which supports
syntax highlighting and automatic indentation for most programming languages.
The submit button runs the code against a series of
test cases. Once the test results are received, they are presented
directly in the browser, giving the student immediate feedback on the
correctness of the submission.
CloudCoder captures the code's standard output and
standard error streams to help students with debugging.

@@

CloudCoder supports two types of exercises:  {\em function-based}
exercises and {\em whole-program} exercises.  Function-based exercises are
very similar to CodingBat exercises: the student is required to write
a function (or method, depending on the programming language) that takes some number of parameter values,
does a computation, and returns a result value.
Students are given a description of
what the function should do, and a couple of example inputs with
expected results.  Writing a new function-based exercise
requires a description, some skeleton code (typically the method
header, although exercises can provide as much or as little code as
desired), and a series of inputs and their corresponding expected results.
Test cases can be marked as ``secret'' so that students don't see their
inputs or expected results, to discourage students from writing a special case to
recognize each input.

Whole-program exercises read input data from
the standard input stream, perform a computation, and write output to the standard output stream.
Expected output is checked by matching a regular expression against the
program's standard output stream.
Whole-program exercises are useful for two main reasons.  First,
they can be assigned early in the course, before functions/methods
have been introduced.  Second, they give the exercise author a
way of checking expected output that is more powerful than matching a return
value against a literal.  The main disadvantage is
the difficulty of devising a
regular expression that is precise enough to distinguish correct output
from incorrect output, but forgiving enough to allow minor deviations
in whitespace, capitalization, and formatting of output.

@@

CloudCoder executes Java, Python, and Ruby code relatively securely using
a Java Security Manager.
(CloudCoder uses Jython\cite{jython} and JRuby\cite{jruby}
to execute Python and Ruby submissions, allowing us to take advantage of
the Java Security Manager for those languages.)
CloudCoder's execution of C/C++ code is somewhat less secure, although 
in the near future
we hope to take advantage of a better sandboxing mechanism,
possibly based on the work of
David Malan for Harvard's online CS50 course\cite{Malan:2013:CSS:2445196.2445242}.
However, because thus far the only users of
CloudCoder have been instructors and students,
the weak sandboxing has not been a significant concern.

@@

As the student works on an exercises, CloudCoder logs all of the
student's edits, and stores all submissions and test outcomes.
Code edits are fine-grained---close to the level of keystroke
logging.
We hope that the data collected by CloudCoder will provide
useful insights to faculty and researchers about
exercises (their difficulty, which knowledge components they develop)
and how students work on them (success patterns, error patterns, etc.)
% combination of exercises and data will give faculty
%and researchers
%insights into the difficulty level of different exercises, keep a
%record of common error patterns, and enable the development of
%innovative data analyses.

One issue is how to manage the data collected.  
Recent work on web-scale data collection by the
BlueJ team\cite{Utting:2012:WDG:2361276.2361278} lays out a number of important
questions about programming data, such as who should have access to the data, and what
mechanism should be used to access the data.  Sanders
et.\ al.\cite{Sanders:2008:DSE:1404520.1404534} reviewed a number of
the challenges involved in sharing empirical education data.

@@

Anonymized data that has been cleared by the proper Institutional
Review Boards is contributed to the Datashop, which
makes the data available to other researchers under whatever
restrictions were put in place by the overseeing IRBs.  The Datashop also
provides a suite of analysis and visualization tools for making
sense of the ITS data.

@@

We recognize that ease of installation is a critical factor for
adoption of a new pedagogical technology for busy
faculty who may only have a few minutes to evaluate a new system.  We
are committed to making the process
of getting up and running as simple as possible.

For evaluation purposes, we run a demo server
(\url{https://cloudcoder.org/demo}) on which any instructor may request
an account.  This provides an easy way for instructors to evaluate
CloudCoder without installing the software.

@@

To date, each institution that has adopted CloudCoder in production has
provided Linux servers that we configured for them.  This has made
adoption very easy, but is not a model that is likely to scale well
to larger numbers of institutions.  We recently received a grant
of credits for Amazon Web Services\cite{aws} from Amazon,
which we plan to use to directly host CloudCoder for interested
institutions.  We are also investigating easier ways for
users to self-host, such as enabling CloudCoder to run on application
hosting services such as Google App Engine\cite{gae}.
%  We have asked Amazon for Amazon Web Services credits to
%provide CloudCoder ``in the cloud'' to adopting institutions, and we
%hope to add support for Google App Engine this summer.  Ease of
%adoption is critical for CloudCoder.

@@

{\bf How can programming exercises be integrated most effectively into a course?}
Programming exercises can be used for many purposes: self-test questions
to accompany readings and video lectures, in-class assessment (possibly
in the context of peer instruction), extra practice for students who
are struggling.  We would like to have specific, evidence-based recommendations
that we can give to instructors who are considering using programming
exercises in their courses to guide them in using exercises in an effective manner.
